---
title: "Instrumental Variables Analysis of Randomized Experiments with One-Sided Noncompliance"
author: 
  - name: Joon-Ho Lee (joonho@berkeley.edu)
  - name: Avi Feller
  - name: Sophia Rabe-Hesketh
date: "February 18, 2019"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: yes
    theme: readable
    highlight: haddock
    css: styles.css
    fig_caption: yes
  tufte::tufte_html:
    number_sections: yes
    toc: true
    toc_depth: 2
  pdf_document:
    toc: yes
    toc_depth: '3'
bibliography: bibliography.bib
---

```{r basic_setup, include=FALSE}
# Set working directory
setwd("~/Bayesian-causal-inference/Case_study_2")

# Set RMarkdown options
knitr::opts_chunk$set(echo = TRUE, warnings = FALSE, message = FALSE)

# Call libraries
library(rstan)
library(rstanarm)
library(bayesplot)
library(tidyverse)
library(gridExtra)
library(kableExtra)

# Set Stan options
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
Sys.setenv(LOCAL_CPPFLAGS = '-march=native')
```


# Introduction




# The Sommer-Zeger vitamin A supplement data

We illustrate the methods presented in this document using the data from a randomized experiment to study the effect of vitamin A supplements on infant mortaility in Indonesia. The data was previously analyzed in @sommer1991estimating and is displayed in the following table:


Table: [Table 1] Sommer-Zeger vitamin A supplement data

| Compliance <br/> Type | Assignment <br/> $Z_{i}$ | Vitamin supplements <br/> $W_{i}^{\text{obs}}$ | Survival <br/> $Y_{i}^{\text{obs}}$ | Number of units <br/> ($N$ = 23,682) |
|--------:|--------:|--------:|--------:|--------:|
| never-taker | 1 | 0 | 0 | 34 |
| never-taker | 1 | 0 | 1 | 2,385 |
| complier | 1 | 1 | 0 | 12 |
| complier | 1 | 1 | 1 | 9,663 |
| complier or never-taker | 0 | 0 | 0 | 74 |
| complier or never-taker | 0 | 0 | 1 | 11,514 |
| *(not observed)* | 0 | 1 | 0 | 0 |
| *(not observed)* | 0 | 1 | 1 | 0 |


In this experiment, villages in Indonesia were randomly assigned to receive or not to receive vitamin A supplements. This *assignment* to the supplements is denoted by $Z_{i} \in \{0, 1 \}$. Some of the individuals in village assigned to the treatment group, however, did not receive the vitamin supplements, while none of the individuals assigned to the control group received the supplements. The *receipt* of the supplements is considered the treatment of main interest and denoted by $W_{i}^{\text{obs}} \in \{0, 1 \}$. Noncompliance in this setting is *one-sided* because all units assigned to the control condition complied with this assignment. The binary outcome variable $Y_{i}^{\text{obs}}$ indicates survival of an infant.          

The eight different combinations of three observed binary variables, $Z_{i}$, $W_{i}^{\text{obs}}$, and $Y_{i}^{\text{obs}}$, are possible in principle. Because all units assigned to the control condition complied with the assignment, however, two combinations with $Z_{i} = 0$ and $W_{i}^{\text{obs}} = 1$ were not observed in the sample. The **R** code that follows loads the dataset based on the counts of the six combinations of observed variables shown in Table. 1. 


```{r import}
# Load data: Sommer-Zeger example from Imbens and Rubin (1997)
z <- c(rep(1, 34 + 2385 + 12 + 9663), 
       rep(0, 74 + 11514))

w <- c(rep(0, 34 + 2385), 
       rep(1, 12 + 9663), 
       rep(0, 74 + 11514))

y <- c(rep(0, 34),
       rep(1, 2385),
       rep(0, 12),
       rep(1, 9663),
       rep(0, 74),
       rep(1, 11514))
```




# Causal estimands when confronted with one-sided noncompliance

Let us first define potential outcomes to fit the instrumental variable settings. For unit $i$, the observed treatment status $W_{i}^{\text{obs}}$ is

$$
W_{i}^{\text{obs}} = W_{i}(Z_{i}) = \begin{cases}
W_{i}(0) & \text{if $Z = 0$}, \\
W_{i}(1) & \text{if $Z = 1$}.
\end{cases}
$$

$W_{i}(0)$ is the treatment that unit $i$ would receive if assinged to the control ($Z_{i} = 0$). $W_{i}(1)$ is the treatment that unit $i$ would receive if assinged to the treatment ($Z_{i} = 1$). The pair of potential responses to treatment assignment for unit $i$, $(W_{i}(0), W_{i}(1))$ describes the compliance behavior. For unit $i$, 

$$
G_{i} = \begin{cases}
c & \text{(unit $i$ is a complier),} & \text{if $W_{i}(z) = z$, for $z = 0, 1$},  \\
n & \text{(unit $i$ is a never-taker),} & \text{if $W_{i}(z) = 0$, for $z = 0, 1$},  \\
a & \text{(unit $i$ is an always-taker),} & \text{if $W_{i}(z) = 1$, for $z = 0, 1$},  \\
d & \text{(unit $i$ is a defier),} & \text{if $W_{i}(z) = 1-z$, for $z = 0, 1$}. 
\end{cases}
$$
We denote the proportions of the four compliance types in the population as $\pi_{g} = \Pr(G_{i} = g)$ for $g \in \{c, n, a, d\}$. The compliance behavior $G_{i}$ is a *latent pre-treatment variable* because it is not fully observed and unaffected by the treatments. 


In the one-sided noncompliance case, all units assigned to the control group complied with the assignment, that is, $W_{i}(0) = 0$ for all units. Thus, the *monotonicity* assumption ($W_{i}(1) \geq W_{i}(0)$) is automatically satisfied and rules out the presence of both always-takers ($a$) and defiers ($d$) in the super-population [@imbens2015causal].      


For the primary outcomes, we define potential outcomes, $Y_{i}(z, w)$, the outcome observed if unit $i$ were assigned treatment $z$ and received treatment $w$. The observed outcome for unit $i$ equals

$$
Y_{i}^{\text{obs}} = Y_{i}(Z_{i}, W_{i}(Z_{i})) = \begin{cases}
Y_{i}(0, 0), & \text{if $Z_{i} = 0, W_{i}^{\text{obs}} = 0$}, \\ 
Y_{i}(0, 1), & \text{if $Z_{i} = 0, W_{i}^{\text{obs}} = 1$}, \\
Y_{i}(1, 0), & \text{if $Z_{i} = 1, W_{i}^{\text{obs}} = 0$}, \\
Y_{i}(1, 1), & \text{if $Z_{i} = 1, W_{i}^{\text{obs}} = 1$}.
\end{cases}
$$


Since both always-takers ($a$) and defiers ($d$) are not present in the one-sided noncompliance setting, we can infer the compliance type for all units with $Z_{i} = 1$. The units with ($Z_{i} = 1, W_{i}^{\text{obs}} = 0$) must be *never-takers*, and the units with ($Z_{i} = 1, W_{i}^{\text{obs}} = 1$) must be *compliers*. For units with $Z_{i} = 0$, however, we could not infer their compliance types. For a unit with ($Z_{i} = 0, W_{i}^{\text{obs}} = 0$), we know that the unit is absent because $W_{i}(0) = 0$ for all unit $i$. But for a unit with ($Z_{i} = 0, W_{i}^{\text{obs}} = 0$), we can only infer that the unit is a *complier* or a *never-taker* because the observed behavior is consistent with those of both compliance types. 


Now consider the intention-to-treat (ITT) effect, the average effect of treatment assignment $Z_{i}$ on the outcome $Y_{i}$. We can decompose the super-population ITT effect in the one-sided noncompliance case into a weighted average of the subgroup effects by two compliance types (compliers and never-takers):


$$
\begin{equation}
\begin{split}
\text{ITT}_{Y} &= \sum_{g \in \{c, n\}}{\mathbb{E}[Y_{i}(1, W_{i}(1))-Y_{i}(0, W_{i}(0))|G_{i} = g]} \cdot\Pr(G_{i} = g) \\
&= \mathbb{E}[Y_{i}(1, 1)-Y_{i}(0, 0)|G_i = c] \cdot \pi_c + \mathbb{E}[Y_{i}(1, 0)-Y_{i}(0, 0)|G_i = n] \cdot \pi_n \\
&= \text{ITT}_Y^{c} \cdot\pi_c + \text{ITT}_Y^n \cdot \pi_n
\end{split}
\end{equation}
$$


The two subgroup effects on $Y$ by compliance type cannot be estimated directly from the observed data because the latent compliance behavior for units assigned to the control condition ($Z_{i} = 0$) cannot be inferred. But under an additional assumption, *exclusion restrictions*, we can still disentangle the ITT effects by compliance status. Exclusion restrictions capture the notion that there is no effect of assignment $Z_{i}$ on the outcome $Y_i$, in the absence of an effect of the assignment of treatment $Z_i$ on the treatment received $W_i$ [@imbens2015causal]. For all units with $G_i = n$ (never-takers), the exclusion restriction requires $Y_i(1, 0)$ to be equal to $Y_i(0, 0)$. In other words, the potential outcomes are unaffected by the assignment $Z_i$ for never-takers who would not receive the treatments even if assigned to take them.              


Because the super-population average ITT effect for never-takers ($\text{ITT}_Y^n$) is zero under the exclusion restriction, the ITT effect on the primary outcome $\text{ITT}_Y$ can be simplified to 

$$
\begin{equation}
\begin{split}
\text{ITT}_{Y} &= \mathbb{E}[Y_{i}(1, 1)-Y_{i}(0, 0)|G_i = c] \cdot \pi_c \\
&= \mathbb{E}[Y_{i}(1)-Y_{i}(0)|G_i = c] \cdot \pi_c \\
&= \text{ITT}_Y^{c} \cdot\pi_c. 
\end{split}
\end{equation}
$$

Note that we are dropping the $z$ argument in the potential outcomes because $z$ is always equal to $w$ under the exclusion restriction for never-takers. The estimand of our primary interest, the complier average causal effect ($\text{ITT}_Y^{c}$), is now the ratio of the ITT effect of $Z_i$ on the outcome $Y_i$ ($\text{ITT}_{Y}$) and the proportion of compliers in the populaion ($\pi_c$).         




# Model-based analysis in randomized experiments with one-sided noncompliance

Now we develop a model-based, or a simulation approach for estimating causal effects in randomized experiments with one-sided noncompliance, maintaining the exclusion restriction for never-takers. The model-based approach developed by @imbens1997bayesian focuses on each individual's compliance status while the standard Wald or Two-Stage Least Squares estimator are based solely on sample averages. See the chapter 25 of @imbens2015causal for the detailed description of the model-based estimation strategy. 


Incorporating the individual-level information about compliance status requires the use of a full likelihood function [@feller2016compared]. To define the actual observed data likelihood function for parameters, we first partition the set of $N$ units into the subsets of units representing each pattern of missing and observed data. In the one-sided noncompliance setting described above, three possible values for $(Z_i, W_i^{\text{obs}})$ are $(0, 0)$, $(1, 0)$, and $(1, 1)$. We denote the mutually exclusive and collectively exhaustive subsets by $\mathcal{S}(0, 0)$, $\mathcal{S}(1, 0)$, and $\mathcal{S}(1, 1)$. Now the full likelihood function in terms of the observed data is given by


$$
\begin{equation}
\begin{split}
\mathcal{L}_{\text{obs}}(\theta|\mathrm{Z}^{\text{obs}}, \mathrm{W}^{\text{obs}}, \mathrm{Y}^{\text{obs}}) &= \prod_{i \in \mathcal{S}(1, 0)}{\Pr(G_i=n|Z_{i}) \cdot f(Y_i(0)|G_i = n, Z_i, \eta_{n1})} \times \\
& \prod_{i \in \mathcal{S}(1, 1)}{\Pr(G_i=c|Z_{i}) \cdot f(Y_i(1)|G_i = c, Z_i, \eta_{c1})} \times \\
& \prod_{i \in \mathcal{S}(0, 0)}{\{\Pr(G_i=n|Z_{i}) \cdot f(Y_i(0)|G_i = n, Z_i, \eta_{n0})} +\Pr(G_i=c|Z_{i}) \cdot f(Y_i(0)|G_i = c, Z_i, \eta_{c0}) \}.
\end{split}
\end{equation}
$$


With binary outcomes and no defiers and always-takers ($\pi_{a}=\pi_{d}=0$), the complete parameter vector $\theta$ consists of five parameters, $(\pi_{c}, \eta_{c0}, \eta_{c1}, \eta_{n0}, \eta_{n1})$. The probability of a unit being a complier in the population is $\pi_{c} = \Pr(G_i = c|\theta)$ and the probability of never-takers is $\pi_{n} = 1-\pi_{c}$. $\eta_{gz}$ is a probability for the outcome distribution defined by compliance status $g = c, n$ and treatment assignment $z = 0, 1$. Since the outcomes are binary, we need only one parameter $\eta_{gz}$, the probability of survival of an infant for compliance type $g$ when assigned $z$, for modeling each of the four binomial outcome distributions.


For the set $\mathcal{S}(1, 0)$ and $\mathcal{S}(1, 1)$, we can infer that units with these patterns of observed compliance behaviors are never-takers and compliers, respectively. Thus, the likelihood contribution from the $i$th such unit is proportional to the conditional distribution of potential outcomes $f(Y_i(w)|G_i = g, Z_i, \eta_{gz})$ with weights given by the compliance type probability $\Pr(G_i=g|Z_{i})$. But for the set $\mathcal{S}(0, 0)$, the likelihood contribution for units in this set is a mixture of outcome distributions of two compliance types: compliers and never-takers. 


The estimand of primary interest is the super-population complier average causal effect $\eta_{c1}-\eta_{c0}$. But we may also be interested in the superpopulation analygue of $\text{ITT}_Y^n$, $\eta_{n1}-\eta_{n0}$. 




# Analyzing the example data with **Stan**

```{r}
# Collect data into a list format suitable for Stan
stan_data <- list(N = length(y), y = y, z = z, w = w)
```



## **Stan** model with exclusion restriction

```
data {
  int<lower=1> N;
  int<lower=0,upper=1> z[N];   // treatment assigned
  int<lower=0,upper=1> w[N];   // treatment received  
  int<lower=0,upper=1> y[N];   // outcomes  
}
parameters {

  // PRINCIPAL STRATUM OUTCOME MEANS
  real<lower=0,upper=1> eta_c0;
  real<lower=0,upper=1> eta_c1;
  real<lower=0,upper=1> eta_nt;

  // OVERALL PROBABILITY OF BEING A COMPLIER
  real<lower=0,upper=1> pi;
} 
model {
  
  // PRIORS FOR OUTCOME (from Imbens & Rubin (1997))
  eta_c0 ~ beta(2, 2);  
  eta_c1 ~ beta(2, 2);  
  eta_nt ~ beta(2, 2);  

  // PRIORS FOR COMPLIER PROBABILITY
  // implicit prior: pi ~ Unif(0,1)

  // MODELS FOR OUTCOME
  for(n in 1:N){
    
    // Never Takers
    if(z[n] == 1 && w[n] == 0){
      target +=  log(1 - pi) + bernoulli_lpmf(y[n] | eta_nt) ;
    }
    
    // Complier (control) or Never Taker
    else if(z[n] == 0 && w[n] == 0){
      target += log_sum_exp(
        log(1 - pi) + bernoulli_lpmf(y[n] | eta_nt),  // Never taker
        log(pi) + bernoulli_lpmf(y[n] | eta_c0) );    // Complier (control)
    }
    
    // Complier (treated)
    else if(z[n] == 1 && w[n] == 1){
      target += log(pi) + bernoulli_lpmf(y[n] | eta_c1) ; // Complier (treat)
    }
  }
}
```


```{r, results="hide", cache=TRUE}
# Compile and run the stan model
stan_fit_ER <- stan( file = "Model_01_CACE_with_Exclusion_Restriction.stan", 
                     data = stan_data, 
                     iter = 1000, chains = 4)
```


```{r}
# Print output
print(stan_fit_ER, probs = c(0.1, 0.5, 0.9), digits = 3)
```


```{r}
# Extract parameters 
eta_c1 <- rstan::extract(stan_fit_ER, pars = "eta_c1")$eta_c1
eta_c0 <- rstan::extract(stan_fit_ER, pars = "eta_c0")$eta_c0

# Calculate treatment effect (in per-10,000 units)
cace <- (eta_c1 - eta_c0)*10^3

# Plot CACE
hist(cace, main = "ITT for Compliers with Exclusion Restriction", 
     xlab = "ITT for Compliers (per 10,000)", col = "grey", border = "white",
     xlim = c(-4, 10), breaks = 30)

# Plot median CACE
abline(v = median(cace), col = 2)

# plot 90% CI
quantile(cace, probs = c(0.05, 0.95))
abline(v = quantile(cace, probs = c(0.05, 0.95)), lty = 2, col = 2)
```



## **Stan** model without exclusion restriction for never takers

```
data {
  int<lower=1> N;
  int<lower=0,upper=1> z[N];   // treatment assigned
  int<lower=0,upper=1> w[N];   // treatment received  
  int<lower=0,upper=1> y[N];   // outcomes  
}
parameters {

  // PRINCIPAL STRATUM OUTCOME MEANS
  real<lower=0,upper=1> eta_c0;
  real<lower=0,upper=1> eta_c1;
  real<lower=0,upper=1> eta_nt0;
  real<lower=0,upper=1> eta_nt1;
  
  // OVERALL PROBABILITY OF BEING A COMPLIER
  real<lower=0,upper=1> pi;
} 
model {
  
  // PRIORS FOR OUTCOME (from Imbens & Rubin (1997))
  eta_c0 ~ beta(2, 2);  
  eta_c1 ~ beta(2, 2);  
  eta_nt0 ~ beta(2, 2);  
  eta_nt1 ~ beta(2, 2);  

  // PRIORS FOR COMPLIER PROBABILITY
  // implicit prior: pi ~ Unif(0,1)

  // MODELS FOR OUTCOME
  for(n in 1:N){
    
    // Never Takers (treat)
    if(z[n] == 1 && w[n] == 0){
      target += log(1 - pi) + bernoulli_lpmf(y[n] | eta_nt1);
    }
    
    // Complier (control) or Never Taker (control)
    else if(z[n] == 0 && w[n] == 0){
      target +=  log_sum_exp(
        log(1 - pi) + bernoulli_lpmf(y[n] | eta_nt0),  // Never taker
        log(pi) + bernoulli_lpmf(y[n] | eta_c0));      // Complier (control)
    }
    
    // Complier (treated)
    else if(z[n] == 1 && w[n] == 1){
      target += log(pi) + bernoulli_lpmf(y[n] | eta_c1) ;  // Complier (treat)
    }
  }
}
```


```{r, results="hide", cache=TRUE}
# Compile and run the stan model
stan_fit_noER <- stan( file = "Model_02_CACE_without_Exclusion_Restriction.stan", 
                       data = stan_data, 
                       iter = 1000, chains = 4)
```


```{r}
# Print output
print(stan_fit_noER, probs = c(0.1, 0.5, 0.9), digits = 3)
```


```{r}
# Extract parameters 
eta_c1_no_er <- rstan::extract(stan_fit_noER, pars = "eta_c1")$eta_c1
eta_c0_no_er <- rstan::extract(stan_fit_noER, pars = "eta_c0")$eta_c0

eta_nt1 <- rstan::extract(stan_fit_noER, pars = "eta_nt1")$eta_nt1
eta_nt0 <- rstan::extract(stan_fit_noER, pars = "eta_nt0")$eta_nt0

# Calculate treatment effect (in per-10,000 units)
cace_no_er <- (eta_c1_no_er - eta_c0_no_er)*10^3
nace <- (eta_nt1 - eta_nt0)*10^3
```


```{r}
# Plot CACE
hist(cace_no_er, main = "ITT for Compliers without Exclusion Restriction", 
     xlab = "ITT for Compliers (per 10,000)", col = "grey", border = "white",
     xlim = c(-4, 10), breaks = 30)

# Plot median CACE
abline(v = median(cace_no_er), col = 2)

# Plot 90% CI
quantile(cace_no_er, probs = c(0.05, 0.95))
abline(v = quantile(cace_no_er, probs = c(0.05, 0.95)), lty = 2, col = 2)
```


```{r}
# Plot NACE
hist(nace, main = "ITT for Never Takers", 
     xlab = "ITT for Never Takers (per 10,000)", col = "grey", border = "white",
     xlim = c(-25, 40), breaks = 30)

# Plot median CACE
abline(v = median(nace), col = 2)

# Plot 90% CI
quantile(nace, probs = c(0.05, 0.95))
abline(v = quantile(nace, probs = c(0.05, 0.95)), lty = 2, col = 2)


# Plot joint distribution
plot(cace_no_er, nace, type = 'n',
     xlab = "ITT for Compliers", ylab = "ITT for Never Takers",
     xlim = c(-4, 10), ylim = c(-30, 30), bty = 'n')
abline(h = 0)
abline(v = 0)
points(cace_no_er, nace, pch = 20)
```

-  Compare with and without exclusion restriction

```{r}
boxplot(cace, cace_no_er,
        names = c("With Excl. Rest.", "Without Excl. Rest."),
        col = "grey",
        ylab = "ITT for Compliers")
```





# References


